---
title: "Conditional Least Square Estimator on Multivariate form"
author: "Mo Li"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Parameter estimator

Let $\theta$ be a vector containing all parameters in the AOP(1) model. The conditional least squares (CLS) estimator of $\mathbf{\theta}$, denoted by $\mathbf{\hat{\theta}}$  minimizes the sum of squares in
\begin{equation}
\label{eqnSSE}
Q(\mathbf{\theta}):=\sum_{t=1}^{n}\sum_{k=1}^{K} (Y_{t,k}-\mathbb{E}[Y_{t,k}|Y_{\{t-1\}}, \mathbf{\theta}])^2,
\end{equation}
\noindent where $\mathbb{E}[Y_{t,k}|Y_{\{t-1\}},\theta]$ is the conditional expectation of $Y_{t,k}$ given the past values $Y_{\{t-1\}}= \{Y_{t-1}, \ldots, Y_{1}\}$. It is the best mean squared error (MSE) predictor of $Y_{t,k}$ from $Y_{\{t-1\}}$, and can be obtained by
\begin{flalign}\label{g_form}
  \mathbb{E}[Y_{t,k}|Y_{\{t-1\}},\mathbf{\theta}] \notag &= P(Y_{t}=k|Y_{\{t-1\}}, \mathbf{\theta}) \\ 
                       \notag &= P(Z_{t}\in B_{k}|Z_{\{t-1\}}\in B_{Y_{\{t-1\}}})\\
                       &= \frac{P(Z_{t}\in B_{k} \cap Z_{t-1}\in B_{y_{t-1}} \cap ... \cap Z_{1}\in B_{y_{1}})}{P(Z_{t-1}\in B_{y_{t-1}} \cap ... \cap Z_{1}\in B_{y_{1}})}
\end{flalign}
The evaluation of this conditional expectation requires calculating a $t$-dimensional Gaussian integral, which can be very difficult and time-consuming, even for moderate $t$. Note that the latent process $\{Z_t\}$ is Gaussian, so the best MSE predictor of $Z_t$ is equivalent to its best linear predictor. Although the categorized variable $Y_t$ is not Gaussian, the linear prediction of $Y_t$ would still be a reasonably good approximation of $\mathbb{E}[Y_{t,k}|Y_{\{t-1\}},\mathbf{\theta}]$. 

Here, we consider $\mathbb{E}[Y_{t,k}|Y_{t-1},\mathbf{\theta}]$, the conditional expectation of $Y_{t,k}$ given only the most recent past value, $Y_{t-1}$, to approximate $\mathbb{E}[Y_{t,k}|Y_{\{t-1\}},\mathbf{\theta}]$ in the CLS formula.

### data simulation function

```{r code1}
rm(list = ls())

options("scipen"=10) # show all digits

library(mvtnorm)
library(Rcpp)
library(RcppArmadillo)
library(foreach)
library(doMC)

##### data simulation
ClipSimulation = function(ci, theta, rho, K, Ts, DesignX, seed=NULL){
  
  if(length(ci) != K-2){stop("Number of cut points and categories NOT match!!!")}
  if(length(theta) != NCOL(DesignX)){stop("Number of Design Matrix columns and coefficients NOT match!!!")}
  
  len_par = length(ci) + length(theta) + length(rho)
  
  if(!is.null(seed)){set.seed(seed)}
  mst = DesignX%*%theta
  et = arima.sim(n=Ts, list(ar=rho), sd=sqrt(1-rho^2)) # sd argument is for WN
  Z = et + mst
  
  ci2 = c(0, ci) # c1 = 0
  clip = function(X=NULL){return(length(which(X>ci2))+1)}
  X_hour = sapply(X=as.vector(Z), FUN = clip)
  
  X_hour_wide = matrix(0, nrow=Ts, ncol=K)
  for (h in 1:Ts) {X_hour_wide[h,X_hour[h]] = 1}
  
  res = list(Z, X_hour, X_hour_wide, DesignX)
  names(res) = c("Z", "X_hour", "X_hour_wide", "DesignX")
  
  return(res)
}
```

### Conditional least square for multivariate form objective function

```{r code2}
AopCLSW = function(par, Xl, Xw, DesignX, lag=1){
  
  Ts = length(Xl)
  K = NCOL(Xw)
  
  num_par = length(par)
  
  seg = c(-Inf, 0, par[1:(K-2)], Inf)
  theta = par[(K-1):(num_par-1)]
  rho = par[num_par]
  
  # mean vector
  mst = DesignX%*%theta
  CondEX = matrix(0, nrow=Ts, ncol=K)
  
  # t = 1 (using marginal expectation)
  tmpprob = pnorm(seg[2:(K+1)]-mst[1]) - pnorm(seg[1:K]-mst[1])
  CondEX[1,] = tmpprob
  
  # t = 2 (using conditonal expectation with lag=1)
  corr_nom = rho^abs(matrix(1:2 - 1, nrow = 2, ncol = 2, byrow = TRUE) - (1:2 - 1))
  diag(corr_nom) = 1
  prevX = Xl[1]
  denom = pnorm(seg[prevX+1]-mst[1]) - pnorm(seg[prevX]-mst[1])
  mu_nom = mst[2:1]
  for(k in 1:K){
    a_nom = seg[c(k, prevX)]
    b_nom = seg[c(k, prevX)+1]
    nom = pmvnorm(lower=a_nom, upper=b_nom, mean=mu_nom, corr=corr_nom)
    CondEX[2,k] = nom/denom
  }
  
  # t > 2 (using conditonal expectation with lag)
  for(t in 3:Ts){
    if(t<=lag){
      # 2 < t <= lag
      corr_nom = rho^abs(matrix(1:t-1, nrow=t, ncol=t, byrow=TRUE) - (1:t - 1))
      diag(corr_nom) = 1
      corr_denom = rho^abs(matrix(1:(t-1)-1, nrow=t-1, ncol=t-1, byrow = TRUE) - (1:(t-1) - 1))
      diag(corr_denom) = 1
      
      prevID = 1:(t-1)
      prevX = Xl[prevID]
      a_denom = seg[prevX]
      b_denom = seg[prevX+1]
      mu_denom = mst[prevID]
      denom = pmvnorm(lower=a_denom, upper=b_denom, mean=mu_denom, corr=corr_denom)
      
      currID = 1:t
      mu_nom = mst[currID]
      
      for(k in 1:K){
        a_nom = seg[c(k, prevX)]
        b_nom = seg[c(k, prevX)+1]
        nom = pmvnorm(lower=a_nom, upper=b_nom, mean=mu_nom, corr=corr_nom)
        CondEX[t,k] = nom/denom
      }
    }else{
      # t > lag (using conditional expectation)
      corr_nom = rho^abs(matrix(1:(lag+1) - 1, nrow = (lag+1), ncol = (lag+1), byrow = TRUE) - (1:(lag+1) - 1))
      diag(corr_nom) = 1
      corr_denom = rho^abs(matrix(1:lag - 1, nrow = lag, ncol = lag, byrow = TRUE) - (1:lag - 1))
      diag(corr_denom) = 1
      prevID = (t-1):(t-lag)
      prevX = Xl[prevID]
      a_denom = seg[prevX]
      b_denom = seg[prevX+1]
      mu_denom = mst[prevID]
      if(lag<2){
        denom = pnorm(b_denom-mu_denom) - pnorm(a_denom-mu_denom)
      }else{
        denom = pmvnorm(lower=a_denom, upper=b_denom, mean=mu_denom, corr=corr_denom)  
      }
      
      currID = t:(t-lag)
      mu_nom = mst[currID]
      for(k in 1:K){
        a_nom = seg[c(k, prevX)]
        b_nom = seg[c(k, prevX)+1]
        nom = pmvnorm(lower=a_nom, upper=b_nom, mean=mu_nom, corr=corr_nom)
        CondEX[t,k] = nom/denom
      }
    }
  }
  
  Q = sum((Xw - CondEX)^2)
  
  # cat("\n ===============")
  # cat("\n par =", par)
  # cat("\n Q =", Q)
  return(Q)
}
```


### Conditional least square for univariate form objective function

```{r code3}
AopCLSL = function(par, Xl, DesignX){
  
  Ts = length(Xl)
  K = nlevels(factor(Xl))
  
  num_par = length(par)
  
  seg = c(-Inf, 0, par[1:(K-2)], Inf)
  theta = par[(K-1):(num_par-1)]
  rho = par[num_par]
  
  # mean vector
  mst = DesignX%*%theta
  CondEX = rep(0, Ts)
  
  # t = 1 (using marginal expectation)
  tmpprob = pnorm(seg[2:(K+1)]-mst[1]) - pnorm(seg[1:K]-mst[1])
  CondEX[1] = sum(tmpprob*(1:K))
  
  # t > 2 (using conditional expectation) !!! only lag 1 now !!!
  corr_nom = rho^abs(matrix(1:2 - 1, nrow = 2, ncol = 2, byrow = TRUE) - (1:2 - 1))
  diag(corr_nom) = 1
  for(t in 2:Ts){
    prevX = Xl[t-1]
    denom = pnorm(seg[prevX+1]-mst[t-1]) - pnorm(seg[prevX]-mst[t-1])
    mu_nom = mst[t:(t-1)]
    for(k in 1:K){
      a_nom = seg[c(k, prevX)]
      b_nom = seg[c(k, prevX)+1]
      nom = pmvnorm(lower=a_nom, upper=b_nom, mean=mu_nom, corr=corr_nom)
      CondEX[t] = CondEX[t] + k*nom/denom
    }
  }
  
  Q = sum((Xl - CondEX)^2)
  
  # cat("\n par =", par)
  return(Q)
}
```

### Single simulation investigation

Starting from simplest case, we have $K=3$ categories and the parameter settings are in the below.

```{r code4}
myseed = 12345

##### Main scripts Start from here
Ts = 500

K = 3
ciT = 0.8615
thetaT = 0.43075
names(thetaT) = c("beta0")
rhoT = 0.5
parT = c(ciT, thetaT, rhoT)

DesignXT = matrix(1, nrow=Ts, ncol=1)
colnames(DesignXT) = "Intercept"

# data simulation
resSim = ClipSimulation(ci=ciT, theta=thetaT, rho=rhoT, K=K, Ts=Ts, DesignX=DesignXT, seed=myseed)
Z = resSim$Z
X_hour = resSim$X_hour
X_hour_wide = resSim$X_hour_wide
DesignXT = resSim$DesignX

#------------------- Parameter Estimation   ------------------------#
## initial value calculation
cout = summary(as.factor(X_hour))
ci_initial = as.vector(qnorm(cumsum(cout/sum(cout))))[1:(K-1)]
p <- which(!is.finite(ci_initial))
if(length(p)!=0){
  if(p==1){
    ci_initial[p] <- -6
  }else{
    ci_initial[p] <-  6
  }
}
phi_initial = pacf(X_hour, plot = F)$acf[1]
par_initial = c(ci_initial[2:length(ci_initial)] - ci_initial[1], -ci_initial[1], 
                    rep(0, NCOL(DesignXT)-1), phi_initial)
## parameter constraints
constrLSE = matrix(0, nrow=K, ncol=length(par_initial))
constrLSE[1,1] = 1
# for(ii in 1:(K-3)){constrLSE[ii+1,c(ii, ii+1)] <- c(-1,1)} # ci
constrLSE[(K-1):K,length(par_initial)] = c(1, -1)
constrLSE_ci = c(rep(0,K-2), -1, -1)

## optimization comparison
OptimW = constrOptim(par_initial, f=AopCLSW,
                     method = "Nelder-Mead", ui=constrLSE,
                     ci=constrLSE_ci, hessian=F,
                     control= list(reltol=1e-05),
                     Xl=X_hour, Xw=X_hour_wide, DesignX=DesignXT)
OptimW

OptimL = constrOptim(par_initial, f=AopCLSL,
                     method = "Nelder-Mead", ui=constrLSE,
                     ci=constrLSE_ci, hessian=F,
                     control= list(reltol=1e-05),
                     Xl=X_hour, DesignX=DesignXT)
OptimL

```

Multiple simulations

```{r code5}
nSim = 200
resL = matrix(0, nrow=nSim, ncol=3)
resWlag1 = resWlag2 = resL
resL_convg = matrix(0, nrow=nSim, ncol=2)
resWlag1_convg = resWlag2_convg = resL_convg
for(iii in 1:nSim){

  # data simulation
  resSim = ClipSimulation(ci=ciT, theta=thetaT, rho=rhoT, K=K, Ts=Ts, DesignX=DesignXT, seed=myseed+iii)
  Z = resSim$Z
  X_hour = resSim$X_hour
  X_hour_wide = resSim$X_hour_wide

  #------------------- Parameter Estimation   ------------------------#
  ## initial value calculation
  cout = summary(as.factor(X_hour))
  ci_initial = as.vector(qnorm(cumsum(cout/sum(cout))))[1:(K-1)]
  p <- which(!is.finite(ci_initial))
  if(length(p)!=0){
    if(p==1){
      ci_initial[p] <- -6
    }else{
      ci_initial[p] <-  6
    }
  }
  phi_initial = pacf(X_hour, plot = F)$acf[1]
  par_initial = c(ci_initial[2:length(ci_initial)] - ci_initial[1], -ci_initial[1],
                  rep(0, NCOL(DesignXT)-1), phi_initial)
  ## parameter constraints
  constrLSE = matrix(0, nrow=K, ncol=length(par_initial))
  constrLSE[1,1] = 1
  # for(ii in 1:(K-3)){constrLSE[ii+1,c(ii, ii+1)] <- c(-1,1)} # ci
  constrLSE[(K-1):K,length(par_initial)] = c(1, -1)
  constrLSE_ci = c(rep(0,K-2), -1, -1)

  ## optimization
  OptimW = constrOptim(par_initial, f=AopCLSW,
                       method = "Nelder-Mead", ui=constrLSE,
                       ci=constrLSE_ci, hessian=F,
                       control= list(reltol=1e-05),
                       Xl=X_hour, Xw=X_hour_wide, DesignX=DesignXT, lag=1)
  resWlag1[iii,] = OptimW$par
  resWlag1_convg[iii,] = c(OptimW$convergence, OptimW$counts[1])
  
  OptimW = constrOptim(par_initial, f=AopCLSW,
                       method = "Nelder-Mead", ui=constrLSE,
                       ci=constrLSE_ci, hessian=F,
                       control= list(reltol=1e-05),
                       Xl=X_hour, Xw=X_hour_wide, DesignX=DesignXT, lag=2)
  resWlag2[iii,] = OptimW$par
  resWlag2_convg[iii,] = c(OptimW$convergence, OptimW$counts[1])
  
  OptimL = constrOptim(par_initial, f=AopCLSL,
                     method = "Nelder-Mead", ui=constrLSE,
                     ci=constrLSE_ci, hessian=F,
                     control= list(reltol=1e-05),
                     Xl=X_hour, DesignX=DesignXT)
  resL[iii,] = OptimL$par
}
```

Summary of the comparison
```{r code6}
mySummary = matrix(NA, nrow=10, ncol=3)
colnames(mySummary) = c("c2", "Intercept", "rho")
rownames(mySummary) = c("TRUE", 
                        "UEst", "UBias", "Usd",
                        "MEstLag1", "MBiasLag1", "MsdLag1",
                        "MEstLag2", "MBiasLag2", "MsdLag2")
mySummary[1,] = parT
mySummary[2,] = colMeans(resL)
mySummary[3,] = colMeans(resL) - parT
mySummary[4,] = apply(resL, 2, sd)
mySummary[5,] = colMeans(resWlag1)
mySummary[6,] = colMeans(resWlag1) - parT
mySummary[7,] = apply(resWlag1, 2, sd)
mySummary[8,] = colMeans(resWlag2)
mySummary[9,] = colMeans(resWlag2) - parT
mySummary[10,] = apply(resWlag2, 2, sd)
mySummary
```

Histogram of the comparison
```{r code7}
myXLowB = apply(resL, 2, min) - 0.1
myXUppB = apply(resL, 2, max) + 0.1
par(mfrow=c(2,3))
hist(resL[,1], xlim = c(myXLowB[1], myXUppB[1]))
hist(resL[,2], xlim = c(myXLowB[2], myXUppB[2]))
hist(resL[,3], xlim = c(myXLowB[3], myXUppB[3]))

hist(resWlag1[,1], xlim = c(myXLowB[1], myXUppB[1]))
hist(resWlag1[,2], xlim = c(myXLowB[2], myXUppB[2]))
hist(resWlag1[,3], xlim = c(myXLowB[3], myXUppB[3]))

hist(resWlag2[,1], xlim = c(myXLowB[1], myXUppB[1]))
hist(resWlag2[,2], xlim = c(myXLowB[2], myXUppB[2]))
hist(resWlag2[,3], xlim = c(myXLowB[3], myXUppB[3]))
```
Convergence check
```{r code8}
# convergence counts
sum(1*(resWlag1_convg[,1] !=0))
sum(1*(resWlag2_convg[,1] !=0))

# average number function evaluation
mean(resWlag1_convg[,2])
mean(resWlag2_convg[,2])
```