---
title: 'CLSE: Comparison with Varin2006 and LiLu2022'
author: "Mo Li"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Parameter estimator

Let $\theta$ be a vector containing all parameters in the AOP(1) model. The conditional least squares (CLS) estimator of $\mathbf{\theta}$, denoted by $\mathbf{\hat{\theta}}$  minimizes the sum of squares in
\begin{equation}
\label{eqnSSE}
Q(\mathbf{\theta}):=\sum_{t=1}^{n}\sum_{k=1}^{K} (Y_{t,k}-\mathbb{E}[Y_{t,k}|Y_{\{t-1\}}, \mathbf{\theta}])^2,
\end{equation}
\noindent where $\mathbb{E}[Y_{t,k}|Y_{\{t-1\}},\theta]$ is the conditional expectation of $Y_{t,k}$ given the past values $Y_{\{t-1\}}= \{Y_{t-1}, \ldots, Y_{1}\}$. It is the best mean squared error (MSE) predictor of $Y_{t,k}$ from $Y_{\{t-1\}}$, and can be obtained by
\begin{flalign}\label{g_form}
  \mathbb{E}[Y_{t,k}|Y_{\{t-1\}},\mathbf{\theta}] \notag &= P(Y_{t}=k|Y_{\{t-1\}}, \mathbf{\theta}) \\ 
                       \notag &= P(Z_{t}\in B_{k}|Z_{\{t-1\}}\in B_{Y_{\{t-1\}}})\\
                       &= \frac{P(Z_{t}\in B_{k} \cap Z_{t-1}\in B_{y_{t-1}} \cap ... \cap Z_{1}\in B_{y_{1}})}{P(Z_{t-1}\in B_{y_{t-1}} \cap ... \cap Z_{1}\in B_{y_{1}})}
\end{flalign}
The evaluation of this conditional expectation requires calculating a $t$-dimensional Gaussian integral, which can be very difficult and time-consuming, even for moderate $t$. Note that the latent process $\{Z_t\}$ is Gaussian, so the best MSE predictor of $Z_t$ is equivalent to its best linear predictor. Although the categorized variable $Y_t$ is not Gaussian, the linear prediction of $Y_t$ would still be a reasonably good approximation of $\mathbb{E}[Y_{t,k}|Y_{\{t-1\}},\mathbf{\theta}]$. 

Here, we consider $\mathbb{E}[Y_{t,k}|Y_{t-1},\mathbf{\theta}]$, the conditional expectation of $Y_{t,k}$ given only the most recent past value, $Y_{t-1}$, to approximate $\mathbb{E}[Y_{t,k}|Y_{\{t-1\}},\mathbf{\theta}]$ in the CLS formula.

### data simulation function

```{r code1}
rm(list = ls())

options("scipen"=10) # show all digits

library(mvtnorm)
library(Rcpp)
library(RcppArmadillo)
library(foreach)
library(doMC)

sourceCpp("../tools/tmpVarin2006.cpp")

##### data simulation
Varin2006Sim <- function(ci, theta, rho, Ts, K, myseed=NULL){
  
  Ts <- Ts + 100 # first 100 observations will be dropped
  len_par <- length(par)
  
  ci   <- c(0, ci)
  
  if(!is.null(myseed)){set.seed(myseed)}
  
  B0 <- rep(1, Ts)
  B1 <- rnorm(Ts, mean=-1,    sd=1)
  B2 <- rnorm(Ts, mean=-0.25, sd=sqrt(0.0324))
  DesignXT <- cbind(B0, B1, B2)
  mu <- DesignXT%*%theta
  
  yt <- rep(0, Ts)
  yt[1] <- 0
  eps <- rnorm(Ts, mean=0, sd=1)
  for(t in 2:(length(yt)-1)){
    yt[t] <- mu[t] + rho*yt[t-1] + eps[t]
  }
  
  clip <- function(X=NULL){return(length(which(X>ci))+1)}
  
  # vector form
  X_hour <- sapply(X=as.vector(yt), FUN = clip)
  # matrix form
  X_hour_wide <- matrix(0, nrow=Ts, ncol=K)
  for (h in 1:Ts) {X_hour_wide[h,X_hour[h]] <- 1}
  
  # delete frist 100 observations and list outputs
  res_sim <- list(DesignXT=DesignXT[101:Ts,], yt=yt[101:Ts], 
                  X_hour=X_hour[101:Ts], X_hour_wide=X_hour_wide[101:Ts,])
  return(res_sim)
}
```

### Single simulation investigation

```{r code2}
myseed = 12345

Ts = 500

# Model: Varion and Vidoni (2006)
K = 7
ciT = c(1.2, 2.2, 3.1, 4.1, 5.3)
thetaT = c(2.9, -0.6,  9)
names(thetaT) = c("beta0", "beta1", "beta2")
rhoT = 0.5

parT = c(ciT, thetaT, rhoT)

res_Varin2006Sim = Varin2006Sim(ci=ciT, theta=thetaT, rho=rhoT, Ts=Ts, K=K, 
                                myseed=12345)
res_sim <- list(Z           = res_Varin2006Sim$yt, 
                X_hour      = res_Varin2006Sim$X_hour,
                X_hour_wide = res_Varin2006Sim$X_hour_wide)

DesignXT <- res_Varin2006Sim$DesignXT
X_hour <- res_sim$X_hour
X_hour_wide <- res_sim$X_hour_wide


#------------------- Parameter Estimation   ------------------------#
## initial value calculation
cout = summary(as.factor(X_hour))
ci_initial = as.vector(qnorm(cumsum(cout/sum(cout))))[1:(K-1)]
p <- which(!is.finite(ci_initial))
if(length(p)!=0){
  if(p==1){
    ci_initial[p] <- -6
  }else{
    ci_initial[p] <-  6
  }
}
phi_initial = pacf(X_hour, plot = F)$acf[1]
par_initial = c(ci_initial[2:length(ci_initial)] - ci_initial[1], -ci_initial[1], 
                rep(0, NCOL(DesignXT)-1), phi_initial)
## parameter constraints
constrLSE = matrix(0, nrow=K, ncol=length(par_initial))
constrLSE[1,1] = 1
for(ii in 1:(K-3)){constrLSE[ii+1,c(ii, ii+1)] <- c(-1,1)} # ci
constrLSE[(K-1):K,length(par_initial)] = c(1, -1)
constrLSE_ci = c(rep(0,K-2), -1, -1)


## optimization comparison

Optim2 = constrOptim(par_initial, f=LSE_cpp1,
                     method = "Nelder-Mead", ui=constrLSE,
                     ci=constrLSE_ci, hessian=F,
                     control= list(reltol=1e-04),
                     X=X_hour_wide, DesignX=DesignXT)
Optim2

Optim3 = constrOptim(par_initial, f=LSE_cpp3,
                     method = "Nelder-Mead", ui=constrLSE,
                     ci=constrLSE_ci, hessian=F,
                     control= list(reltol=1e-04),
                     Xl=X_hour, X=X_hour_wide, DesignX=DesignXT)
Optim3
```
